---
title: "Exploratory Analysis"
date: "2023-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
lc <- read.csv("datasets/processed/lc.csv")
```

## First moment descriptive statistics

Plot Canada Lynx within BC

```{r}
plot(decimalLatitude ~ decimalLongitude,
     pch = 16,
     col = "#046C9A",
     data = lc)
```

Visualizing the observation window

```{r}
library(spatstat)
library(sf)
load("datasets/raw/BC_Covariates.Rda")
# Create a SpatialPolygons object for the window
bc_window_sf <- st_as_sf(DATA$Window)
bc_window_owin <- as.owin(bc_window_sf)
# Visualize the window
plot(bc_window_owin, main = "Observation Window")
```

Plot observations in BC window

```{r}
# Convert to a ppp object
lc_ppp <- ppp(x = lc$decimalLongitude, # X coordinates
              y = lc$decimalLatitude, # Y coordinates
              window = bc_window_owin, # Observation window
              )

# Access the rejected points
# attr(lc_ppp, "rejects")

#  Remove rejected/out of window points
lc_ppp <- as.ppp(lc_ppp)

# Plot the ppp object
plot(lc_ppp, 
     pch = 16,
     cex = 0.5,
     cols = "#046C9A",
     main = "Marked Canada Lynx")
```

```{r}
#Get units
unitname(lc_ppp)
```


```{r}
summary(lc_ppp)
```

### Intensity

```{r}
#Split into a 3 by 3 quadrat and count points
Q <- quadratcount(lc_ppp,
                  nx = 10,
                  ny = 10)
#Plot the output
plot(lc_ppp,
     pch = 16,
     cex = 0.5,
     cols = "#046C9A",
     main = "Lynx locations")

plot(Q, cex = 2, col = "red", add = T)
```

```{r}
#Plot the output Note the use of image = TRUE
plot(intensity(Q, image = T),
     main = "Lynx intensity")

plot(lc_ppp, pch = 16, cex = 0.6, cols = "white", add = T)

plot(lc_ppp, pch = 16, cex = 0.5, cols = "black", add = T)
```

Clearly, the assumption of homogeneity is not appropriate for this dataset as the lynx tend to be clustered in certain areas of the study site, whereas others have no trees at all.

Quadrat test of homogeneity

```{r}
quadrat.test(Q)
```

The small p-value suggests that there is a significant deviation from homogeneity.

### Relationships with covariates

We are usually interested in determining whether the intensity depends on a covariate(s). One simple approach to check for a relationship between inhomogeneous $\lambda(u)$ and a spatial covariate $Z(u)$ is via quadrat counting.

```{r}
# Create 5 classes with equal width
elev_classes <- cut(DATA$Elevation, breaks = 5, labels = c("low", "low-medium", "medium", "high-medium", "high"))
table(elev_classes[lc_ppp])
```

## Second Moment Descriptives

### K-Function

```{r}
# Estimate the k- function
k_lc <- Kest(lc_ppp)
# visualise the results
plot (k_lc, main = "Homogeneous K-Function", lwd = 2)
```

Here we can see that theoretical k-function $k_{pois}(r)$ deviates from other k-function corrections indicates clustering but these estimates assume homogeneity.

```{r}
# Bootstrapped CIs
# rank = 1 means the max and min
# values will be used for CI
E_lc <- envelope(finpines , Kest , rank = 1, nsim = 19, fix.n = T)
# visualise the results
plot (E_lc , main = "Homogeneous K-function")
```

Now we have evidence that suggests significant clustering, but these estimates assume homogeneity.

Relaxing homogeneity assumption.

```{r}
#Estimate intensity
lambda_lc <- density(lc_ppp, bw.ppl)
Kinhom_lc <- Kinhom(lc_ppp, lambda_lc)
# Estimate a strictly positive density
lambda_lc_pos <- density(lc_ppp, sigma=bw.ppl, positive=TRUE)
# Simulation envelope (with points drawn from the estimated intensity)
E_lc_inhom <- envelope(lc_ppp, 
                        Kinhom, 
                        simulate = expression(rpoispp(lambda_lc_pos)), 
                        correction="border", 
                        rank = 1, 
                        nsim = 19, 
                        fix.n = TRUE)
# visualise the results
plot(E_lc_inhom, xlim = c(0,320000), main = "Inhomogeneous K-function", lwd = 2)
```

When correcting for inhomogeneity, the clustering is not as strong homogeneous k-function. Significant clustering appears to exist in and around 0 to 125000 units, 200000 to 217000, 250000 to 300000.

### Pair Correlation Function

```{r}
#Simulation envelope (with points drawn from the estimated intensity)
pcf_lc_inhom <- envelope(lc_ppp, 
                          pcfinhom, 
                          simulate = expression(rpoispp(lambda_lc_pos)), 
                          rank = 1, 
                          nsim = 19)

# visualise the results
plot(pcf_lc_inhom, main = "Inhomogeneous g-function")
```

Seems like clustering occurs for ca. 15000.
